What should I notice? Evaluating memorability of events for abductive inference.

### 2021-08-17 paper.tex
Starting to take JL's corrections into account

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{url}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage[ruled,linesnumbered]{algorithm2e}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{
  What should I notice? Evaluating memorability of events to generate surprising causal candidates.
}
\author{
  \IEEEauthorblockN{
  \IEEEauthorrefmark{1}\IEEEauthorrefmark{2} Étienne Houzé,
  \IEEEauthorrefmark{1} Jean-Louis Dessalles,
  \IEEEauthorrefmark{1} Ada Diaconescu,
  \IEEEauthorrefmark{2} David Menga
  }

@@

As the number of connected devices and sensors grows, so does the amount of
produced data. As a consequence, the ability to filter out a few events from past recordings
as ``memorable'' becomes prime. While this task is often overlooked by many
statistical approaches, which instead benefit from large quantity of
information, it is part of the learning process for humans
\cite{dehaene_how_2020}. Faced with new situations, humans are prone to infer
previously seen unusual events as possible causes, and carry out further testing
to assess the existence of a causal relation.

For instance, it is possible, for humans without any prior knowledge about
physics or electrical engineering, to infer a memorable thunderstorm as a
possible cause for a general black-out just by using the memorability of
the former. While this kind of abductive inference (i.e. inference to the cause
of an observation \cite{magnani_abduction_2011}) can yield many
false-positives, it can be considered in situations where the lack of knowledge
from previous occurrences penalizes other approaches.

@@

This kind of abduction seems however hard to translate into machine, as the difficulty
comes from different factors. First, events can be of different nature, and not
directly comparable. In fact, even for specific systems such as smart homes,
events range from device removal to a presence detection or an unusually high
temperature. How can one can then assess which ones are more memorable?
Furthermore, even for comparable events, different characteristics can be put
forward to argue for the most memorable: is a record-high temperature 47 days
ago more memorable than the small deviation recorded just 3 minutes ago? To this
day, no current system proposes to use all this information from various event
types from different devices to compute a unified metrics of ``memorability''.

@@

To tackle this challenge, we use as a starting point the observation that while
all events, regardless of their characteristics or nature, can be uniquely
described using a combination of qualifiers, the most memorable ones are likely
to require less words to be described. Think, for instance, of ``last year's
hottest day'' and ``the 182th day of 7 years ago''. By evaluating the lenght of
each description, taking into account both the complexity of concept words (a
time ranking, a temperature ranking), and the arguments used in both
descriptions (the hottest, the 182th, 7), we can assign each event a complexity
score. Memorable events, as they stand out, would then differ from their
neighbors by being much simpler (or more complex).

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{figures/filters}
  \caption{An example of retrieving an event through successive predicative
    filters. From the base memory (yellow), successive filters select the events
    satisfying their predicates (gray arrows). The operations yields successive
    subsets of the memory (yellow), until a final singleton memory is reach,
    meaning that the successive filters retrieve a unique event from the memory.
    The predicates used for this retrieval path and their arguments,
    $\pi_{1}, \pi_{2}, \pi_{3}, k_{1}, k_{2}, k_{3}$, can be used as a retrieval
    program. Its length gives an upper bound to the description complexity of
    the event $e_{h}$: ``last year's hottest day''.}
  \label{fig:filters}
\end{figure}

To programmatically achieve this computation, we rely on principles from
Algorithmic Information Theory, which provides formal definitions of complexity
and simplicity of objects, which appear to be consistent with human perception
\cite{li_introduction_2008, dessalles2011coincidences, delahaye_numerical_2012}.
 Figure \ref{fig:filters} shows our solution applied to the example description
 ``last year's hottest day''. We assume that
a memory $\mathcal{M}$ is a unordered set of various distinct objects, upon
which can be applied boolean predicates $\pi_{k}$ to describe them. These predicates thus
provide filters $f_{\pi, k}$ that can be applied to the memory, selecting all events
satisfying predicate $\pi_{k}$. By applying successive
filters, we narrow down the size of the memory, until a single event remains.
All the successive filters used to come to this singleton memory constitutes a
\emph{retrieval path}, whose length, in bits, gives an upper bound of the
description complexity of the retrieved event.

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{url}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage[ruled,linesnumbered]{algorithm2e}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}
\title{
  What should I notice? Evaluating memorability of events for abductive inference.
}
\author{
  \IEEEauthorblockN{
  \IEEEauthorrefmark{1}\IEEEauthorrefmark{2} Étienne Houzé,
  \IEEEauthorrefmark{1} Jean-Louis Dessalles,
  \IEEEauthorrefmark{1} Ada Diaconescu,
  \IEEEauthorrefmark{2} David Menga
  }

@@

% As the number of connected devices and sensors grows, so does the amount of
% produced data. As a consequence, the ability to filter out a few events from past recordings
% as ``memorable'' becomes prime. While this task is often overlooked by many
% statistical approaches, which instead benefit from large quantity of
% information, it is part of the learning process for humans
% \cite{dehaene_how_2020}. Faced with new situations, humans are prone to infer
% previously seen unusual events as possible causes, and carry out further testing
% to assess the existence of a causal relation.

As a user has just turned on the TV in her all-equipped living-room, the
lights dim and the window blinds lower. Intrigued by this behavior, she
quickly infers that both light dimming and the blinds closing occurred as a
consequence of the TV being turned on. How did she come to this conclusion?
By performing \emph{abductive inference} \cite{magnani_abduction_2011}, which
is a key element of human's ability to understand the world: from the
observed consequences, infer the possible causes.

In this example, there are mainly three possibilities to come to the
conclusion. If the user knows how the smart living-room system works, if she
knows the set of applied rules or parameters, she has access to a causal
knowledge she can use for the abduction. Otherwise, if she has no knowledge
about the system but the behavior occurs frequently, examining past
correlations will reveal that the television frequently causes the blinds to
close and the lights to dim. Else, if this is the first instance of the TV
being turned on in this living-room, there is no previous occurrence upon
which the user may rely to draw conclusions. However, she remains able to
infer the TV as a possible cause for the observed reaction: by noticing the
TV as a memorable recent event (since it is its first occurrence), she
proposes it as a possible cause for the questioned phenomenon. This example
shows how humans are able to use distinct methods to perform abductive tasks
and infer new knowledge. While abduction of the first two kinds can be
automated using knowledge bases and statistical approaches, this paper
proposes an approach to evaluate the memorability of events, allowing
machines to perform ``memorability-based'' abduction.

 % For instance, it is
% possible, for humans without any prior knowledge about
% physics or electrical engineering, to infer a memorable thunderstorm as a
% possible cause for a general black-out just by using the memorability of
% the former. While this kind of abductive inference (i.e. inference to the cause
% of an observation \cite{magnani_abduction_2011}) can yield many
% false-positives, it can be considered in situations where the lack of knowledge
% from previous occurrences penalizes other approaches.

@@

The difficulty of implementing this kind of abduction for machines comes from
several factors. First, events can be of different nature, and not directly
comparable. In fact, even for specific systems such as smart homes, events
range from device removal to a presence detection or an unusually high
temperature. How can one can then assess which ones are more memorable?
Furthermore, even for comparable events, different characteristics can be put
forward to argue for the most memorable: is a record-high temperature 47 days
ago more memorable than the small deviation recorded just 3 minutes ago? To
this day, no current system proposes to use all this information from various
event types from different devices to compute a unified metrics of
``memorability''.

@@

% To tackle this challenge, we use as a starting point the observation that while
% all events, regardless of their characteristics or nature, can be uniquely
% described using a combination of qualifiers, the most memorable ones are likely
% to require less words to be described. Think, for instance, of ``last year's
% hottest day'' and ``the 182th day of 7 years ago''. By evaluating the lenght of
% each description, taking into account both the complexity of concept words (a
% time ranking, a temperature ranking), and the arguments used in both
% descriptions (the hottest, the 182th, 7), we can assign each event a complexity
% score. Memorable events, as they stand out, would then differ from their
% neighbors by being much simpler (or more complex).

To tackle this challenge, we rely on the following observation: while all
events, regardless of their characteristics or nature, can be uniquely
described using a combination of qualifiers, the most memorable ones are
likely to require less words to be described. Think, for instance, of how
``last year's hottest day'' and ``the 182nd day from 7 years ago'': the
former seems simpler to describe. By evaluating the length of each
description, taking into account both the complexity of concept words (a date
of occurrence, a temperature ranking), and the arguments (the hottest, the
182ndThis
kind of inference, named \emph{abductive inference} \cite
{magnani_abduction_2011}, is a key element of humans's ability to understand the world., 7) used in both descriptions, we can assign each event a \emph{description complexity}
score. Memorable events, as they stand out, would then differ from their
neighbors and require simpler and less numerous qualifiers to be
unambiguously described. As such, they will stand out as simpler.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\linewidth]{figures/filters}
  \caption{An example of retrieving an event through successive predicative
   filters. From the base memory (yellow), successive filters select the
   events satisfying the associated predicate (grey arrow). For example,
   filter $f_1$ selects events from last year, i.e. which satisfy the
   predicate $\mathtt{date}(event, 1\_year)$. In this case, successively
   applying filters $f_1$, $f_2$ and $f_3$ yields a unique event $e_h$, last
   year's hottest day. The complexity of this event can then be upper bounded
   by the complexity of the three filters, since they give a unambiguous way
   to describe the event in the memory.}
   % their predicates (gray arrows). The
   % operations yields successive subsets of the memory (yellow), until a final
   % singleton memory is reach, meaning that the successive filters retrieve a
   % unique event from the memory. The predicates used for this retrieval path
   % and their arguments, $\pi_{1}, \pi_{2}, \pi_{3}, k_{1}, k_{2}, k_{3}$, can
   % be used as a retrieval program. Its length gives an upper bound to the
   % description complexity of the event $e_{h}$: ``last year's hottest
   % day''.}
  \label{fig:filters}
\end{figure}

For machines to use description complexity to distinguish memorable events from others, we need a formal canvas and computation methods which coincide with human intuition . To do so, we rely on principles from Algorithmic Information Theory, which also appear to be consistent with the human perception of complexity\cite{li_introduction_2008,dessalles2011coincidences,delahaye_numerical_2012}. Our method is as follows: we
consider events as being unordered elements from a base \emph{memory}. To
reproduce the language features applicable to events, we use \emph
{predicates}, i.e. functions assigning a boolean value to an event and
eventual arguments. For instance, the predicate $\mathtt{date}
(\cdot, 1\_year)$ assigns \emph{true} to events that occurred last year. Selecting
all events, from the memory, that satisfy a given predicate corresponds to
a \emph{filter} operation, and yields another memory, a subset of the former.
This filtering operation can then be repeated, each iteration, selecting
fewer events, until a singleton memory is reached, meaning that the sequence
of predicates can \emph{retrieve} unambiguously the unique remaining event.
The description complexity of this event can thus be upper-bounded by the
number of bits required to describe the filters used in the retrieval
process. Figure \ref{fig:filters} illustrates this evaluation for the event
$e_h$: ``last year's hottest day''.

% To programmatically achieve this computation, we rely on principles from
% Algorithmic Information Theory, which provides formal definitions of complexity
% and simplicity of objects, which appear to be consistent with human perception
% \cite{li_introduction_2008, dessalles2011coincidences, delahaye_numerical_2012}.
%  Figure \ref{fig:filters} shows our solution applied to the example description
%  ``last year's hottest day''. We assume that
% a memory $\mathcal{M}$ is a unordered set of various distinct objects, upon
% which can be applied boolean predicates $\pi_{k}$ to describe them. These predicates thus
% provide filters $f_{\pi, k}$ that can be applied to the memory, selecting all events
% satisfying predicate $\pi_{k}$. By applying successive
% filters, we narrow down the size of the memory, until a single event remains.
% All the successive filters used to come to this singleton memory constitutes a
% \emph{retrieval path}, whose length, in bits, gives an upper bound of the
% description complexity of the retrieved event.
